Experiments on Parallel Training of Deep Neural Network using Model Averaging

### 2015-07-05 paper.tex
prepare to submit to arXiv

% Template for ICIP-2013 paper; to be used with:
%          spconf.sty  - ICASSP/ICIP LaTeX style file, and
%          IEEEbib.bst - IEEE bibliography style file.
% --------------------------------------------------------------------------
\documentclass{article}
\usepackage{spconf,amsmath,graphicx}

@@

% Title.
% ------
\title{Investigation on Parallel Training of Deep Neural Network using Model Averaging}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this work we apply model averaging to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging across nodes is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes,
which allows us to perform model averaging frequently without losing much time on communication.
We investigate the effectiveness of Natrual Gradient for Stochasitc Gradient Descent (NG-SGD) in the model averaging
framework, and compare two learning rate schedule strategies for model update. It is shown that NG-SGD leads to better
convergence in model averaging framework. On the 300h Swithboard dataset, a 9.3 times speedup is achieved using 16 GPUs
and 17 times speedup using 32 GPUs with limited decoding accuracy loss. 
\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure of DNNs.

@@

Optimality for parallelization of DNN training was analyzed in \cite{seide2014parallelizability}, and based on the analysis, 
a gradient quantization approach (1-bit SGD) was proposed to minimize communication cost \cite{seide20141}. It shows that 1 bit
can effectively reduce the amount of data exchanged in an MPI framework, and a 10 times speed-up using 40 GPUs.

@@

\subsection{Switchboard Results}
Fig.~\ref{fig:scaling} shows the scaling factor and speedup plot for model averaging experiments. As is shown in the graph,
a speedup of 17 could be achieved when 32 GPUs are used.
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{scaling.png}
  \caption{Scaling factor and speedup factor v.s. number of gpus}
  \label{fig:scaling}
\end{figure}
Table~\ref{tab:wer} shows the main decoding results for DNNs trained using different number of GPUs. In general, decoding
results of DNNs trained model averaging degrades 0.3\~0.4 WER, depending on the number of GPUs used.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
    \hline
     Nodes  & 1    & 2    & 4   & 8      & 16     & 32\\
    \hline
    SWB     & 14.7 & 14.9 & --  & 15.1   & 15.1   & 15.2\\
    CallHM  & 26.8 & 26.8 & --  & 27.4   & 27.0   & 27.1 \\
    \hline
  \end{tabular}
  \caption{Comparison of WERs using different number of GPUs}
  \label{tab:wer}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes          &  1     & 32     &   1    & 32\\
    \hline
    random init     & --    & 16.4   &  --    & 28.8 \\
    \hline
    RBM pretraining & 14.7  & 15.2   & 26.8   & 27.1 \\
    \hline
  \end{tabular}
  \caption{Comparing RBM pretraining with random initialization}
  \label{tab:init}
\end{table}
As we can see in the table, random initialization is worse than RBM pretraining based DNN by xx in single GPU case.
In model averaging setup, random initialization becomes even worse -- a xx point more degradation on WER.

@@

\subsection{Minibatch Size}
Table~\ref{tab:mbsize} compares two different minibatch size in model averaging setup. It is shown that.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}    & \multicolumn{2}{c|}{CallHome} & Speedup\\
    \hline
    nodes       &  1    &   16    &    1     &   16   &    \\
    \hline
    256         &  --      &      &  --      &        & \\
    \hline
    1024        & 9.32     & 15.1 & 27.0     &        & \\
    \hline
  \end{tabular}
  \caption{Comparing different minibatch size}
  \label{tab:mbsize}
\end{table}

@@

As is shown in Table~\ref{tab:lr}, exponential learning rate gives slight better decoding results than Newbob.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes       &  1      & 16    &  1    & 16 \\
    \hline
    Newbob      & 14.9    & 15.4  & 26.6  & 27.2 \\
    \hline
    exponential & 14.7    & 15.1  & 26.8  & 27.0 \\
    \hline
  \end{tabular}
  \caption{Comparing learning rate schedule}
  \label{tab:lr}
\end{table}

\subsection{Online NG-SGD Matters}
Table~\ref{tab:ngsgd} compares plain SGD with NG-SGD in model averaging mode, and it shows NG-SGD with 
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
     Nodes  &  1    & 16        &   1   &  16 \\
    \hline
    SGD     & 14.9    & 16.3    &  26.9 & 28.3 \\
    \hline
    NG-SGD  & 14.7    & 15.1    &  26.8 & 27.0  \\
    \hline
  \end{tabular}
  \caption{Comparing NG-SGD and naive SGD}
  \label{tab:ngsgd}
\end{table}

\section{Conclusion and Future Work}
In this paper, we show that neural network training can be efficiently speeded up using model averaging. 
on a 300h Switchboard dataset, a 9.3x / 17x speedup could be achieved using 16 / 32 GPUs respectively, 
with limited decoding accuracy loss. We also show that model averaging benefits a lot from NG-SGD and RBM based pretraining.
Preliminary experiments on minibatch size, averaging frequency and learning rate schedules are also presented.

>>>>

% Template for ICIP-2013 paper; to be used with:
%          spconf.sty  - ICASSP/ICIP LaTeX style file, and
%          IEEEbib.bst - IEEE bibliography style file.
% --------------------------------------------------------------------------
\documentclass{article}
\usepackage{spconf,amsmath,graphicx}
\usepackage{slashbox,pict2e}

@@

% Title.
% ------
\title{Experiments on Parallel Training of Deep Neural Network using Model Averaging}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this work we apply model averaging to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging across nodes is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes,
which allows us to perform model averaging frequently without losing much time on communication.
We investigate the effectiveness of Natrual Gradient Stochasitc Gradient Descent (NG-SGD) and Restricted Boltzmann 
Machine (RBM) pretraining for parallel training in model-averaging framework, and explore the best setups in 
term of different learning rate schedules, averaging frequencies and minibatch sizes. It is shown that NG-SGD 
and RBM pretraining benefits parameter-averaging based model training. On the 300h Swithboard dataset, a 9.3 times 
speedup is achieved using 16 GPUs and 17 times speedup using 32 GPUs with limited decoding accuracy loss. 
\footnote{This work is not submitted to 
peer-review conferences because the authors think it needs more investigation. The authors are in lack of 
resources to perform further exploration. However, we welcome any comments and suggestions.}
\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network, natural gradient
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure.

@@

Optimality for parallelization of DNN training was analyzed in \cite{seide2014parallelizability}, and based on the analysis, 
a gradient quantization approach (1-bit SGD) was proposed to minimize communication cost \cite{seide20141}. It shows that 1 bit
quantization can effectively reduce data exchange in an MPI framework, and a 10 times speed-up is achieved using 40 GPUs.

@@

\subsection{Switchboard Results}
Fig.~\ref{fig:scaling} shows the scaling factor and speedup plot for model averaging experiments. As is shown in the graph,
a speedup of 17 could be achieved when 32 GPUs are used.
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{scaling.png}
  \caption{Scaling factor and speedup factor v.s. number of gpus}
  \label{fig:scaling}
\end{figure}
Table~\ref{tab:wer} shows the main decoding results for DNNs trained using different number of GPUs. In general, decoding
results of DNNs trained model averaging degrades 0.3\~0.4 WER, depending on the number of GPUs used.
\begin{table}
  \centering
  \begin{tabular}{|c|c|c|c|c|c|c|}
    \hline
    \backslashbox{Data}{Nodes}  & 1    & 2    & 4   & 8      & 16     & 32\\
    \hline
    SWB     & 14.7 & -- & --  & 15.1   & 15.1   & 15.2\\
    CallHM  & 26.8 & -- & --  & 27.4   & 27.0   & 27.1 \\
    \hline
    SWB     & 16.1  & -- & -- & 16.4  & 16.2  & 16.4 \\
    SWB2P3  & 21.0  & -- & -- & 21.8  & 21.7  & 21.7 \\
    SWB-Cell & 27.4 & -- & -- & 27.3  & 27.4  & 27.8 \\
    \hline
  \end{tabular}
  \caption{Comparison of WERs using different number of GPUs}
  \label{tab:wer}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes          &  1     & 32     &   1    & 32\\
    \hline
    random init     & 15.6  & 16.4   & 27.4   & 28.8 \\
    \hline
    RBM pretraining & 14.7  & 15.2   & 26.8   & 27.1 \\
    \hline
  \end{tabular}
  \caption{Comparing RBM pretraining with random initialization}
  \label{tab:init}
\end{table}
As we can see in the table, random initialization is worse than DNN with RBM pretraining by 0.9/0.6 in single GPU case.
While in model averaging setup, random initialization becomes even worse -- 0.3/0.9 point more degradation on WER.

@@

\subsection{Minibatch Size}
Table~\ref{tab:mbsize} compares two different minibatch size in model averaging setup.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}    & \multicolumn{2}{c|}{CallHome} & Speedup\\
    \hline
    nodes       &  1    &   16    &    1     &   16   &    \\
    \hline
    256         & 15.3     & 15.6 & 26.8     &  27.3   & -- \\
    \hline
    1024        & 14.7     & 15.1 & 26.8     &  27.0   & 9.32\\
    \hline
  \end{tabular}
  \caption{Comparing different minibatch size}
  \label{tab:mbsize}
\end{table}

@@

As is shown in Table~\ref{tab:lr}, these two learning rate scheduling methods give similar decoding results. However, 
exponential learning rate might need more tuning since it requires a initial learning rate, a final learning rate and
predefined number of epochs to train.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
    Nodes       &  1      & 16    &  1    & 16 \\
    \hline
    Newbob      & 14.9    & 15.4  & 26.6  & 27.2 \\
    \hline
    exponential & 14.7    & 15.1  & 26.8  & 27.0 \\
    \hline
  \end{tabular}
  \caption{Comparing learning rate schedule}
  \label{tab:lr}
\end{table}

\subsection{Online NG-SGD Matters}
Table~\ref{tab:ngsgd} compares plain SGD with NG-SGD in model averaging mode, and it shows NG-SGD is crucial to 
model training with parameter-averaging.
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}  & \multicolumn{2}{c}{CallHome} \\
    \hline
     Nodes  &  1    & 16        &   1   &  16 \\
    \hline
    SGD     & 14.9    & 16.3    &  26.9 & 28.3 \\
    \hline
    NG-SGD  & 14.7    & 15.1    &  26.8 & 27.0  \\
    \hline
  \end{tabular}
  \caption{Comparing NG-SGD and naive SGD}
  \label{tab:ngsgd}
\end{table}

\section{Conclusion and Future Work}
In this work, we show that neural network training can be efficiently speeded up using model averaging. 
on a 300h Switchboard dataset, a 9.3x / 17x speedup could be achieved using 16 / 32 GPUs respectively, 
with limited decoding accuracy loss. We also show that model averaging benefits a lot from NG-SGD and RBM based pretraining.
Preliminary experiments on minibatch size, averaging frequency and learning rate schedules are also presented.
